---
title: "Chasing Billions: The Global Unicorn Landscape"
subtitle: "Funding, Valuation, and Growth Dynamics of Billion-Dollar Startups"
author: "Himangshu Chetia"
date: "2026-01-16"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
---

## The Business Problem
A Venture Capital (VC) firm wants to identify "efficient" unicorns (reaching $1B in <5 years) outside Silicon Valley to find underserved, high-growth markets.

## Analytical Pipeline (Tech Stack)
* **Excel:** Data entry, validation, and complex string manipulation to convert string values like `$8B` into integer values.
* **SQL:** Structured querying and aggregation of the 1000+ company dataset.
* **R:** Converting data from wide to long format, statistical analysis and R Markdown reporting.
* **Tableau:** [Link to Dashboard] Interactive visualization of investor networks.

## Key Performance Indicators (KPIs)
1. **Speed to Unicorn:** Years from foundation to $1B+
2. **Capital Efficiency Ratio:** Valuation $\div$ Total Funding.
3. **Elite Investors:** Investors with stakes in the top 10% most efficient startups.
4. **Market Saturation:** Density of unicorns per city/industry.

## Data Integrity & Cleaning
* **Currency Normalization:** Converted mixed-string funding `($B/$M)` into integers for calculation.
* **Investor Parsing:** Created a "long" version of the data (where one company has multiple rows, one for each investor) using R & the `tidyverse()` package.
* **Spatial Correction:** Imputed missing city data for Singapore, Hong Kong and Bahamas entities.
* **Outlier Handling:** Addressed data entry errors (e.g., founded dates listed after join dates).

## Limitations & Assumptions
* **Survival Bias:** Data excludes companies that have already exited via IPO or Acquisition.
* **Founding Date Assumption:** Companies are assumed to be founded on Jan 1st of their founding year for time-series calculations.

## Data Source
* **Dataset:** Unicorn Companies Global Valuations
* **Platform:** Kaggle
* **Author:** Adil Shamim
* **Description:** This dataset provides a comprehensive list of "Unicorn" companies (private companies valued at over $1 billion). It includes key metrics such as current valuation, total funding raised, industry sector, headquarters location (City, Country, Continent), the year the company was founded, and the names of primary investors.
* **Citation:** Shamim, A. (2024). Startup growth and investment data [Data set]. Kaggle. https://www.kaggle.com/datasets/adilshamim8/startup-growth-and-investment-data

# Phase 1: Data Engineering & Pre-processing (Excel/R)

*Goal: Perform heavy-duty aggregation and multi-dimensional filtering.*

## Currency Normalization

Used the following Excel formula to convert shorthand ($8B, $200M) into long-form integers:

```excel
=IF(RIGHT(B2,1)="B", 
    SUBSTITUTE(SUBSTITUTE(B2,"$",""),"B","")*1000000000, 
    IF(RIGHT(B2,1)="M", 
        SUBSTITUTE(SUBSTITUTE(B2,"$",""),"M","")*1000000, 
        SUBSTITUTE(B2,"$","")*1
    )
)
```

## Date Standardization

* Converted all dates into the `YYYY-MM-DD` format

* Identified and fixed outliers in the date columns where `Year Founded` > `Date Joined` using the `DATEDIF` function.

* Imputed missing values (like the Singapore,Hong Kong and Bahamas city gap).

## Data Transformation

### Wide to Long Format

Created a dataset (dataset_long.csv)containing only company and investor columns by converting the wide data into long format using the following R script:

```r
# Load the necessary libraries
library(tidyverse)

# Read the data
df <- read_csv("dataset_long_temp.csv")

# Convert from wide to long
df_long <- df %>%
separate_rows(investors, sep = ",") %>%
mutate(investors = str_trim(investors))

# View result
view(df_long)

# Save the file
readr::write_csv(df_long, "dataset_long.csv")
```
### Geocoding

Created a dataset (geocoded_unicorns.csv) cntaining the Latitude and Longitude of the cities.

# Phase 2: Database Management & SQL Analysis

*Goal: Establish the baseline metrics.*

## The "Efficiency" Query:

Performed the following SQL query to filter the companies that reached Unicorn status within 5 years:

```sql
SELECT company, years_to_unicorn
FROM (
	SELECT
        company,
        YEAR(date_joined) - YEAR(year_founded) AS years_to_unicorn
	FROM dataset
	GROUP BY company, years_to_unicorn
    ) AS yu
WHERE years_to_unicorn <= 5
ORDER BY years_to_unicorn;
```

## Geographic Grouping:

Performed the following query to find the `Average Valuation` per startup per country:

```sql
SELECT
	country,
    avg(valuation) AS average_valuation
FROM dataset
GROUP BY country
ORDER BY average_valuation DESC;
```

## Investor Counting:

Wrote the following query to identify which specific investors appear most frequently in the "Top 10%" of the most efficient startups:

```sql
WITH enriched AS (
    SELECT
        company,
        investors,
        YEAR(date_joined) - YEAR(year_founded) AS years_to_unicorn
    FROM dataset_long
    WHERE YEAR(date_joined) - YEAR(year_founded) <= 5
),
ranked AS (
    SELECT
        company,
        investors,
        years_to_unicorn,
        NTILE(10) OVER (ORDER BY years_to_unicorn) AS decile
    FROM enriched
)

SELECT
    investors,
    COUNT(*) AS companies_invested
FROM ranked
WHERE decile = 1
GROUP BY investors
ORDER BY companies_invested DESC;
```

## Top valued industries:

Identified industries with the highest company valuations.

```sql
SELECT
    industry,
    SUM(valuation) AS total_valuation
FROM dataset
GROUP BY industry;
```

# Phase 3: Statistical Discovery & Advanced EDA (R)

*Goal: Use R to uncover statistical relationships, validate assumptions, and extract non-obvious structure from the global unicorn dataset through rigorous exploratory and inferential analysis.*


## Preaparation

Load the necessary libraries, a function to format large numbers, set the working directory, load the datasets and calculate time to unicorn status.

```{r setup, message = FALSE}
library(tidyverse)
library(scales)
library(knitr)
library(viridis)
library(maps)
library(treemapify)
library(patchwork)

dollar_format_big <- function(x) {
  case_when(
    x >= 1e9 ~ paste0(x / 1e9, "B"),
    x >= 1e6 ~ paste0(x / 1e6, "M"),
    x >= 1e3 ~ paste0(x / 1e3, "K"),
    TRUE ~ as.character(x)
  )
}

setwd(r"(/home/himangshu/Desktop/My_Projects/The_Global_Unicorn_Landscape)")

data <- read_csv("dataset.csv") %>%
  mutate(years_to_unicorn = year(date_joined) - year(year_founded))

data_long <- read_csv("dataset_long.csv")

data_geocoded <- read_csv("geocoded_unicorns.csv")

glimpse(data)
glimpse(data_long)
glimpse(data_geocoded)
```

## Time-based Analysis

### Distribution of Years Taken to Reach Unicorn Status

```{r dist-years-to-unicorn, warning = FALSE}
ggplot(data, aes(x = years_to_unicorn, y = after_stat(density))) +
  geom_histogram(fill = "grey70", colour = "white", bins = 50, alpha = 0.8) +
  geom_density(colour = "#404080") +
  geom_vline(xintercept = mean(data$years_to_unicorn), colour = "#D55E00", linetype = "dashed") +
  geom_vline(xintercept = median(data$years_to_unicorn), colour = "#0072B2", linetype = "dashed") +
  annotate("text", x = 12, y = 0.1, label = "Mean", colour = "red") +
  annotate("text", x = 2, y = 0.1, label = "Median", colour = "blue") +
  labs(
    title = "Distribution of Years to Reach Unicorn Status",
    subtitle = sprintf("Mean: %.2f, Median: %.2f", 
                       mean(data$years_to_unicorn), 
                       median(data$years_to_unicorn)),
    x = "Years to $1B valuation",
    y = "Density"
  ) +
  theme_minimal() +
  xlim(0, 50)
```

### Unicorns Founded Per Year & Joined Per Year

```{r unicorns-founded-joined-per-year}
founded_counts <- data %>%
  mutate(year = year(year_founded)) %>%
  group_by(year) %>%
  summarise(count = n(), .groups = "drop")

joined_counts <- data %>%
  mutate(year = year(date_joined)) %>%
  group_by(year) %>%
  summarise(count = n(), .groups = "drop")

plot_founded <- ggplot(founded_counts, aes(x = year, y = count)) +
  geom_line(linewidth = 1.1, color = "#F8766D") +
  labs(
    title = "Unicorns Founded Per Year",
    x = "Year",
    y = "Number of Unicorns"
  ) +
  theme_minimal(base_size = 13)

plot_joined <- ggplot(joined_counts, aes(x = year, y = count)) +
  geom_line(linewidth = 1.1, color = "#00BFC4") +
  labs(
    title = "Unicorns Joined Per Year",
    x = "Year",
    y = "Number of Unicorns"
  ) +
  theme_minimal(base_size = 13)

plot_founded
plot_joined
```

### Key Findings

* The "Golden Era" (2010–2016): The majority of today's unicorns were founded in this window, with a peak of over 150 companies being founded.

* The 2021 Anomaly: There was a massive, unprecedented spike in companies becoming unicorns in 2021 (>500 companies), likely driven by aggressive venture capital activity post-pandemic.

* Time to Success: It takes an average of 7 years (Mean: 7.01) to reach a $1 billion valuation, though the median is slightly faster at 6 years. Some companies took 25+ years to reach unicorn status, thus pulling the mean towards right.

* The Lag Effect: The sharp drop in "Founding" numbers after 2016 isn't a decline in startups; it simply reflects that companies founded recently haven't had the requisite ~6-7 years growth window to reach unicorn status yet.

* Consistent Correlation: The ~6-year gap between the "Founding Peak" (2015) and the "Unicorn Status Peak" (2021) perfectly aligns with the median growth timeline shown in the distribution chart.


## Analysing Relationships Between Variables.

### Correlation Matrix

```{r correlation-matrix}
data %>%
  mutate(
    year_founded_num = year(year_founded),
    year_joined_num = year(date_joined)
  ) %>%
  select(valuation, funding, years_to_unicorn, year_founded_num, year_joined_num) %>%
  drop_na() %>%
  cor(method = "spearman") %>%
  as.data.frame() %>%
  mutate(var1 = rownames(.)) %>%
  pivot_longer(-var1, names_to = "var2", values_to = "correlation") %>%
  ggplot(aes(var1, var2, fill = correlation)) +
  geom_tile(color = "white", linewidth = 0.6) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  geom_text(aes(label = round(correlation, 2)), size = 4) +
  labs(
    title = "Spearman Correlation Heatmap",
    x = NULL,
    y = NULL,
    fill = "ρ"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### Total Funding vs. Valuation

```{r funding-vs-valuation, message = FALSE, warning = FALSE}
ggplot(data, aes(x = funding, y = valuation, color = industry)) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C0392B") +
  scale_x_log10(labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_log10(labels = label_number(scale_cut = cut_short_scale())) +
  scale_color_viridis_d() +
  labs(title = "Total Funding vs. Valuation",
       x = "Total Funding (USD)", 
       y = "Current Valuation (USD)", 
       caption = "Note: Log scale axes") +
  theme_minimal()
```

### Year Founded vs. Year Joined Unicorn Status

```{r founded-vs-joined}
data %>%
  mutate(
    year_founded_num = year(year_founded),
    year_joined_num = year(date_joined)
  ) %>%
  ggplot(aes(x = year_founded_num, y = year_joined_num)) +
  geom_point(aes(color = years_to_unicorn), alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  scale_color_viridis_c(name = "Years to\nUnicorn") +
  labs(
    title = "Year Founded vs Year Joined Unicorn Status",
    subtitle = "Color indicates time taken to reach unicorn status",
    x = "Year Founded",
    y = "Year Joined"
  ) +
  theme_minimal()
```

### Total Funding vs. Time to Unicorn

```{r funding-vs-time-to-unicorn, message = FALSE, warning = FALSE}
ggplot(data, aes(x = funding, y = years_to_unicorn)) +
  geom_point(alpha = 0.8, color = "#457297ff") +
  geom_smooth(method = "lm", se = FALSE, color = "#C0392B") +
  scale_x_log10(labels = dollar_format_big, breaks = scales::log_breaks(n = 6)) +
  labs(
    title = "Total Funding vs. Years to Unicorn",
    x = "Total Funding (USD)",
    y = "Years to Unicorn",
    caption = "Note: Log scale with LOESS trend"
  ) +
  theme_minimal()
```

### Valuation vs Tine to Unicorn

```{r valuation-vs-time-to-unicorn, message = FALSE, warning = FALSE}
ggplot(data, aes(x = years_to_unicorn, y = valuation, color = continent)) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "#C0392B") +
  scale_y_log10(labels = dollar_format_big) +
  scale_color_viridis_d() +
  labs(
    title = "Valuation vs. Years to Unicorn",
    x = "Years to Unicorn",
    y = "Current Valuation (USD)",
    caption = "Note: Log scale with linear trend"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )
```

### Key Findings

* Money Equals Value (Mostly): There is a moderately strong positive correlation (0.59) between total funding and current valuation. This confirms the intuitive link that companies raising more capital generally achieve higher valuations, though the scatter plot shows significant variance, meaning capital efficiency varies widely by industry.

* Speed is Accelerating: There is a very strong negative correlation (-0.85) between the year a company was founded and the years it took to become a unicorn. This indicates that newer companies are reaching unicorn status much faster than their predecessors.

* The "Speed limit" of Success: The "Valuation vs. Years to Unicorn" chart reveals a flat trend line. Getting to unicorn status faster does not correlate with a higher ultimate valuation. A company that took 2 years to reach $1B is not necessarily more valuable today than one that took 10 years.

* Funding Doesn't Buy Speed: Interestingly, the correlation between total funding and years to unicorn is effectively zero (-0.03). The scatter plot confirms a flat trend line, suggesting that simply pouring more money into a startup does not statistically shorten the time required to reach a $1B valuation.

* The "Wall" of 2021: The "Year Founded vs Year Joined" chart clearly visualizes the "lag" mentioned earlier. The data hits a hard "wall" around 2021-2022 where the red dashed line (representing the present day) cuts off the data. This visualizes why we saw the drop-off in the previous "Founded" charts: companies founded after 2017 simply haven't hit the timeline boundary yet.

## Industry Deep-Dive: `AI` vs `Fintech`

### Summary statistics table
```{r summary-stats-ai-vs-fintech}
data %>%
  filter(industry %in% c("Artificial intelligence", "Fintech")) %>%
  group_by(industry) %>%
  summarise(
    Company_Count = n(),
    Avg_Years_to_Unicorn = mean(years_to_unicorn, na.rm = TRUE),
    Median_Years_to_Unicorn = median(years_to_unicorn, na.rm = TRUE),
    Avg_Funding_USD = mean(funding, na.rm = TRUE),
    Median_Funding_USD = median(funding, na.rm = TRUE)
  ) %>%
  kable(caption = "Statistical Profile: AI vs. Fintech")

comparison_data <- data %>%
  filter(industry %in% c("Artificial intelligence", "Fintech"))

comparison_stats <- comparison_data %>%
  group_by(industry) %>%
  summarise(
    Avg_Funding_USD = mean(funding, na.rm = TRUE),
    Median_Funding_USD = median(funding, na.rm = TRUE)
  )
```

### Comparing growth rate
```{r growth-rate-ai-vs-fintech}
ggplot(comparison_data, aes(industry, years_to_unicorn, fill = industry)) +
  geom_boxplot(alpha = 0.8) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  scale_fill_viridis_d() +
  labs(title = "Distribution of Growth Velocity", 
       subtitle = "Years from Founding to $1B Valuation",
       x = "Industries",
       y = "Years") +
  theme_minimal()
```

### Funding Requirements (Mean Funding)
```{r funding-requirements-ai-vs-fintech}
ggplot(comparison_stats, aes(industry, Avg_Funding_USD, fill = industry)) +
  geom_col() +
  geom_errorbar(aes(ymin = Avg_Funding_USD - Median_Funding_USD,
                    ymax = Avg_Funding_USD + Median_Funding_USD), width = 0.2) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_fill_viridis_d() +
  labs(title = "Average Capital Requirement",
       x = "Industries",
       y = "Average Funding Raised (USD)") +
  theme_minimal()
```

### Key Findings

* Speed to Scale: AI startups reach unicorn status faster (median 5 years) compared to Fintech (median 6 years), showing a tighter, more consistent growth timeline.

* The Funding Divergence:

  * Average: AI is higher ($557M vs. $486M) due to capital-intensive foundational models.

  * Median: AI is lower ($302.5M vs. $370M), indicating the typical AI startup is actually more capital-efficient than Fintech.

* The "Barbell" Effect: The AI sector is split between expensive "Model Builders" (skewing the average up) and efficient "Model Users" (keeping the median low).

* Regulatory Costs: Fintech faces a high "minimum floor" for funding due to licensing, compliance teams, and balance sheet capital requirements.

* Burn Rate Advantage: AI startups save money simply by being faster; the 1-year speed advantage reduces the amount of "survival cash" or runway needed.

* Revenue Scalability: AI offers exponential revenue growth with near-zero marginal costs, whereas Fintech scales linearly based on transaction volume and costs.

## Geographic Analysis

### Unicorn World Map
```{r unicorn-world-map, message = FALSE}
ggplot() +
  geom_polygon(data = map_data("world"), aes(long, lat, group = group), 
               fill = "gray80", color = "white") +
  geom_point(data = data_geocoded, aes(longitude, latitude, size = valuation, 
             color = continent), alpha = 0.6) +
  scale_size_continuous(labels = dollar_format_big, name = "Valuation (USD)") +
  scale_color_viridis_d() +
  labs(title = "Global Distribution of Unicorn Companies", x = "", y = "", 
       color = "Continent") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.margin = margin(t = 10)
  ) +
  guides(
    size = guide_legend(order = 1, nrow = 1),
    color = guide_legend(order = 2, nrow = 1)
  )
```

### Top 10 Unicorn Countries
```{r top-10-unicorn-countries}
data %>%
  count(country) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(reorder(country, n), n, fill = country)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_d() +
  labs(title = "Top 10 Unicorn Countries", x = "Country", y = "Number of Unicorns") +
  theme_minimal()
```

### Top 10 Unicorn Cities
```{r top-10-unicorn-cities}
data %>%
  count(city) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(reorder(city, n), n, fill = city)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_d() +
  labs(title = "Top 10 Unicorn Cities", x = "City", y = "Number of Unicorns") +
  theme_minimal()
```

### Valuation by Country and Industry

```{r treemap-valuation-country-industry}
data %>%
  group_by(country, industry) %>%
  summarise(total_valuation = sum(valuation), .groups = "drop") %>%
  ggplot(aes(area = total_valuation, fill = country, subgroup = country,
             label = paste(industry, dollar_format_big(total_valuation), sep = "\n"))) +
  geom_treemap() +
  geom_treemap_subgroup_border(size = 1) +
  geom_treemap_subgroup_text(place = "centre", grow = TRUE, alpha = 0.5, 
                             colour = "black", fontface = "italic", size = 14) +
  geom_treemap_text(colour = "white", place = "topleft", reflow = TRUE, size = 10) +
  scale_fill_viridis_d() +
  labs(title = "Treemap: Country and Industry") +
  theme_minimal(base_size = 16) +
  theme(legend.position = "none")
```

### Key Findings

* U.S. Dominance: The United States is the overwhelming global leader, hosting approximately 3x as many unicorns as China.

* Top Cities: San Francisco (150+) and New York (~100) lead globally; Palo Alto's presence in the top 10 further highlights the Bay Area's density.

* Asian Powerhouses: China ranks 2nd (led by Beijing, Shanghai, Shenzhen) and India ranks 3rd (anchored by Bengaluru).

* European Hubs: The UK leads in Europe, followed by Germany and France, with London, Berlin, and Paris as key city centers.

* Geographic Concentration: Unicorns are heavily concentrated in the Northern Hemisphere, with significantly lower density in South America, Africa, and Oceania.

* Valuation Drivers by Region:

  * U.S.: Dominated by Fintech ($454B) and Internet Software ($444B).

  * China: Focused on AI ($214B) and E-commerce ($187B).

  * U.K.: Primarily concentrated in Fintech ($147B).

## Industry and Valuation Analysis

### Unicorns by Industry
```{r unicorns-by-industry}
data %>%
  group_by(industry) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  ggplot(aes(x = reorder(industry, count), y = count, fill = industry)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis(discrete = TRUE) +
  labs(
    title = "Number of Unicorns by Industry",
    x = "Industry",
    y = "Number of Unicorns"
  ) +
  theme_minimal()
```

### Total Valuation by Industry
```{r valuation-by-industry}
data %>%
  group_by(industry) %>%
  summarise(total_valuation = sum(valuation)) %>%
  arrange(desc(total_valuation)) %>%
  ggplot(aes(x = reorder(industry, total_valuation), y = total_valuation, fill = industry)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis(discrete = TRUE) +
  scale_y_continuous(labels = dollar_format_big) +
  labs(
    title = "Total Valuation by Industry",
    x = "Industry",
    y = "Total Valuation (USD)"
  ) +
  theme_minimal()
```

### Key Findings

* Dominant Industries: Fintech is the leading industry globally, ranking first in both the number of unicorn companies (>200) and total valuation (~$800B+). Internet Software & Services follows as a strong second in both categories.

* US Sector Strength: The United States holds the largest valuation share across most sectors, driven primarily by Fintech ($454B) and Internet Software & Services ($444B).

* China's AI Lead: While the US dominates overall, China surpasses the US in the Artificial Intelligence sector, with a total valuation of $214B compared to the US's $135B.

* Regional Specializations:

  * China shows a heavy concentration in E-commerce ($187B) and Hardware.

  * The United Kingdom is a major hub for Fintech, which accounts for $147B of its unicorn valuation.

* Valuation Consistency: The top four industries—Fintech, Internet Software, E-commerce, and AI—remain consistent across both company count and total valuation, indicating these are the most mature and high-value sectors in the startup ecosystem.

## Investor Analysis

### Top 10 Investors

```{r top-10-investors}
data_long %>%
  filter(company %in% (data %>% filter(valuation >= quantile(valuation, 0.9)) %>% pull(company))) %>%
  count(investors, name = "count") %>%
  slice_max(count, n = 10) %>%
  ggplot(aes(x = reorder(investors, count), y = count, fill = investors)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_d() +
  labs(
    title = "Top 10 Investors in Most Valuable Unicorns",
    x = "Investor",
    y = "Number of Unicorns"
  ) +
  theme_minimal()
```

## Key Findings
* The distribution of years to reach unicorn status is right-skewed, with a median of 6 years and a mean of 7 years.
* There is a positive correlation between funding and valuation, with a Pearson correlation coefficient of ~0.60 and a Spearman correlation coefficient of ~0.59. Funding accounts for about 36% of the variation in valuation.
* AI companies tend to reach unicorn status faster but require more funding than Fintech companies.
* AI investing favors larger checks with faster scaling, while Fintech offers lower average capital requirements but longer, less predictable paths to unicorn status.

# Phase 4: Visual Storytelling & BI (Tableau)

## Executive Summary (The "TL;DR")
This analysis provides a data-driven perspective on the structural realities of startup growth and capital formation. By examining both the distribution of time required to achieve unicorn status and the variation in average funding across industries, the findings highlight meaningful heterogeneity in growth trajectories and capital intensity. The results suggest that success timelines are not uniform and are strongly influenced by sector-specific dynamics, reinforcing the importance of contextual benchmarks when evaluating startup performance.

While the analysis offers valuable descriptive insights, it is limited by the scope and granularity of the available data. Future work could extend this approach by incorporating longitudinal funding rounds, geographic controls, or macroeconomic conditions to better isolate causal relationships. Overall, this phase establishes a solid empirical foundation for deeper inferential or predictive modeling in subsequent stages of the project.
